---
title: "Interpretation of Data Final Project"
author: "Yash Sinojia A00268852"
date: "05/12/2019"
output: html_document
---

# Importing Data and Libraries

  Various libraries have been loaded to utlilize its diverse functionalities. viridis, hrbrthemes and scales are some color packages to add a layer on ggplot2 visualizations.

```{r}

transport_2011 <- read.csv('transportCSOfile.csv')
headers <- names(transport_2011)

library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(viridis)
library(hrbrthemes)
library(scales)

```

# Data Cleansing

## Refining Headers

```{r}

headers_travelby <- headers[10:22]
  
```

    It is indeed difficult to use such long header names to access data for further analysis.
    
    According to the nature of the column categories of the dataset and the repetitive wordings of headers, first, the headers are divided into three sections.

```{r}

headers_travelby <- str_split_fixed(headers_travelby, "_", 18)[,c(13:16)]
headers_travelby <- paste("Travelby", headers_travelby[,1], headers_travelby[,2], 
                              headers_travelby[,3], headers_travelby[,4], sep = " ")


headers_travel_time <- headers[23:32]
headers_travel_time <- str_split_fixed(headers_travel_time, "_", 17)[,c(15:17)]
headers_travel_time <- paste("Leaving", headers_travel_time[,1], headers_travel_time[,2],
                             headers_travel_time[,3], sep = " ")

headers_journey <- headers[33:40]
headers_journey <- str_split_fixed(headers_journey, "_", 20)[,c(12:20)]
headers_journey <- paste("Journey", headers_journey[,1], headers_journey[,2], headers_journey[,3],
                         headers_journey[,4], headers_journey[,5], headers_journey[,6], 
                         headers_journey[,7], headers_journey[,8], headers_journey[,9])

```
    
    To manipulate the headers, 'str_split_fixed' function from 'stringr' library has been used. Here, the headers have been sliced on the basis of underscores and the unique tailing part of the string has been extracted out of it. 

```{r}

headers[10:22] <- headers_travelby
headers[23:32] <- headers_travel_time
headers[33:40] <- headers_journey

headers[10:40] <- str_split_fixed(headers[10:40], '2011', 2)[,1]
headers <- trimws(headers)
headers <- make.names(headers, unique=TRUE)

```

    The method 'trimws' has been used to remove the lingering spaces at the end of each header string.
    Furthermore, it is essential to remove any spaces in the middle of words for it to be a accessible header. For that, 'make.names' method has been used.

```{r}

names(transport_2011) <- headers
names(transport_2011)

```

## Handling Missing Values

    Any action taken on handling missing values must be framed on the basis of the nature of dataset and the proportion of missing values in it.

    To find the number and location of the missing values, 'is.na' with 'which' to trigger row and coulumn indices of those values.
    
```{r}

which(is.na(transport_2011), arr.ind = TRUE)

```
    
    Here, there are only two null values in the dataset.
    
    One missing value is at the aggregate total column for 'Travel Modes' section. Handled by replacing the value by taking the sum of all the previous values in that particular row section.
    
```{r}

travelby_total = rowSums(transport_2011[10:18])
transport_2011[29,22] <- travelby_total[29]

```
    
    The other null value can't be determined as the previous one. So this value has to be replaced by mean as the measure of centrality of column where the value resides.
    
```{r}

transport_2011[79,30] <- ceiling(mean(transport_2011[,30]))

which(is.na(transport_2011), arr.ind = TRUE)


```


## Handling Outliers

    All the value outside the upper and lower bounds of Interquartile range are outliers. Bounds are formed by quantile function.
    
    If the difference between mean and median is significant, the outlier is replaced by median data, if not, it is replaced by mean.

```{r, echo=FALSE}
knitr::opts_chunk$set(error = FALSE)
```

treat_outliers <- function(df){
  summary(df)
  sd(df)
  boxplot(df, xlab='before')
  boxplot.stats(df)
  if(abs(mean(df) - median(df)) <= 5){
    df[df>(quantile(df,prob=.75)) + (1.5*IQR(df))] <- mean(df)
    df[df<(quantile(df,prob=.25)) - (1.5*IQR(df))] <- mean(df)
    boxplot(df, xlab='after')
  }
  else if(is.na(abs(mean(df) - median(df)))){}
  else{
    df[df>(quantile(df,prob=.75)) + (1.5*IQR(df))] <- median(df)
    df[df<(quantile(df,prob=.25)) - (1.5*IQR(df))] <- median(df)
    boxplot(df, xlab='after')
  }
  return(df)
}

for(i in 10:40){
  transport_2011[,i] <- treat_outliers(transport_2011[,i])
}

```

    Due to treating the outliers, the section 'Total's and 'Comb's may have differed a bit, so rowSum aggregate is calculated and stored back to those columns.

```{r}

transport_2011$Travelby.Total <- rowSums(transport_2011[10:18])
transport_2011$Leaving.Total <- rowSums(transport_2011[23:31])
transport_2011$Journey.Total <- rowSums(transport_2011[33:39])

transport_2011$Travelby.Soft.Modes.Comb <- rowSums(transport_2011[10:11])
transport_2011$Travelby.Public.Transport.Comb <- rowSums(transport_2011[12:13])
transport_2011$Travelby.Private.Transport.Comb <- rowSums(transport_2011[14:17])

```

# Storing the cleaned data

    Now, the data is clean and has to be exported out as a .csv file.

```{r}

write.csv('cleaned_TransportCSOfile.csv')

```

# Answers to Questions

## Answer 1.

   Found by taking the sum of each column in the Travel Mode section of the dataframe.
    
```{r}

df <- as.data.frame(colSums(transport_2011[10:18]))
df <- cbind(rownames(df),df)
names(df)[1] <- 'Travelby'
names(df)[2] <- 'Total'
df[,2] <- as.integer(df[,2])

p<-ggplot(data = df, aes(x=df$Travelby, y=df$Total)) +
  geom_bar(stat="identity", fill = "#56B4E9") +
  geom_text(aes(label=df$Travelby), vjust=-1, size=3.5)  +
  theme_minimal() +
  xlab("Mode of Transports Nationally") +
  ylab("Count")
p

```

    The use of Car as the mode of transportation has been most popular in the nation.


## Answer 2.

    The data has to be filtered out with County column values as 'Wexford'.

```{r}

transport_2011_wexford <- transport_2011[transport_2011$County == 'Wexford',]

df2 <- as.data.frame(colSums(transport_2011_wexford[10:18]))
df2 <- cbind(rownames(df),df)
names(df)[1] <- 'Travelby'
names(df)[2] <- 'Total'
df[,2] <- as.integer(df[,2])

p<-ggplot(data = df, aes(x=df$Travelby, y=df$Total)) +
  geom_bar(stat="identity", fill = "steelblue") +
  geom_text(aes(label=df$Travelby), vjust=-1, size=3.5)  +
  theme_minimal() +
  xlab("Mode of Transports in County Wexford") +
  ylab("Count")
p

```

   The most popular transport in Wexford is Car, same as the national preference.


## Answer 3.

    All the values containing the word 'City' are extracted out in a vector. Which is further used with '%in%' operator to subset out the rows that have those values. Similar operation is performed for 'Non-City' values.
    

```{r}

transport_2011_city <- grep("City", transport_2011$County, value = TRUE)
transport_2011_noncity <- grep("City", transport_2011$County, value = TRUE, invert = TRUE)

transport_2011_citydf <- subset(transport_2011, transport_2011$County %in% transport_2011_city)
transport_2011_noncitydf <- subset(transport_2011, transport_2011$County %in% transport_2011_noncity)

df1 <- as.data.frame(colSums(transport_2011_citydf[10:18]))
df2 <- as.data.frame(colSums(transport_2011_noncitydf[10:18]))
df1 <- cbind(rownames(df1),df1)
df2 <- cbind(rownames(df2),df2)
dfc <- as.data.frame(c(rep('City',9)))
dfnc <- as.data.frame(c(rep('Non-City',9)))
df1 <- cbind(df1, dfc)
df2 <- cbind(df2, dfnc)

names(df1)[1] <- 'Travelby'
names(df1)[2] <- 'Total'
names(df1)[3] <- 'Region'
names(df2)[1] <- 'Travelby'
names(df2)[2] <- 'Total'
names(df2)[3] <- 'Region'
df1[,2] <- as.integer(df1[,2])
df2[,2] <- as.integer(df2[,2])
df <- rbind.data.frame(df1, df2)

```

    Visualizations are juxtaposed with two barplots. It can be noted that Car is most preferred in both region and the distribution is a bit similar except with some categories.

```{r}

ggplot(data = df, aes(fill=df$Region, y=df$Total, x=df$Travelby)) + 
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=df$Travelby), vjust=-1, size=3.5)  +
  scale_fill_viridis(discrete = T, option = "E") +
  ggtitle("Transports in Cities vs. Non-Cities") +
  facet_wrap(~df$Region) +
  theme_classic() +
  xlab("Areas") +
  ylab("Frequency")

```

## Answer 4.

```{r}

df <- cbind.data.frame(c(sum(colSums(transport_2011[c(23:26,29:30)]))/colSums(transport_2011[32]), 
                      1-sum(colSums(transport_2011[c(23:26,29:30)]))/colSums(transport_2011[32])))

names(df)[1] <- 'ratio'


ggplot(data = df, aes(x=" ", y = df$ratio, fill=c('outside_8to9am','total'))) +
  geom_bar(stat="identity", position="identity") +
  coord_polar("y", start=0) +
  geom_text(aes(y = df$ratio, 
                label = percent(df$ratio)), size=5) +
  theme_minimal() +
  scale_fill_brewer(palette="Set2") +
  ylab("Proportion of commuters leaving outside 8 to 9 hr")


```

    Almost half of the total commuters leave outside of 8 - 9 AM rush hour in the morning.


## Answer 5.

```{r}

df <- as.data.frame(c(sum(colSums(transport_2011_wexford[c(36:37)]))/colSums(transport_2011_wexford[40]),
                      1-sum(colSums(transport_2011_wexford[c(36:37)]))/colSums(transport_2011_wexford[40])))
names(df)[1] <- 'ratio'

ggplot(data = df, aes(x=" " , y=df$ratio, fill=c('longer than 45 min','total'))) +
  geom_bar(stat="identity") +
  coord_polar("y", start=0) +
  geom_text(aes(y = df$ratio + c(0, cumsum(df$ratio)[-length(df$ratio)]), 
                label = percent(df$ratio)), size=5) +
  scale_fill_brewer(palette="Blues")+
  theme_minimal() +
  ylab("Ratio of commuters travelling longer than 45 min in County Wexford")
  

```

    Only `sum(colSums(transport_2011_wexford[c(36:37)]))/colSums(transport_2011_wexford[40]*100`% of the commuters travel longer than 45min in County Wexford.

## Answer 6.

```{r}

south_east_other <- transport_2011[transport_2011$NUTS_III == 'South-East' & transport_2011$County != 'Wexford',]
df <- as.data.frame(c(sum(colSums(south_east_other[c(36:37)]))/colSums(south_east_other[40]),
                      1-sum(colSums(south_east_other[c(36:37)]))/colSums(south_east_other[40])))
names(df)[1] <- 'ratio'

ggplot(data = df, aes(x=" " , y=df$ratio, fill=c('longer than 45 min','total'))) +
  geom_bar(stat="identity") +
  coord_polar("y", start=0) +
  geom_text(aes(y = df$ratio + c(0, cumsum(df$ratio)[-length(df$ratio)]), 
                label = percent(df$ratio)), size=5) +
  scale_fill_brewer(palette="Reds")+
  theme_minimal() +
  ylab("Ratio of commuters travelling longer than 45 min in other counties of same NUTS III region")

```

    Only `(sum(colSums(south_east_other[c(36:37)]))/colSums(south_east_other[40])*100`% of the commuters travel longer than 45min in other counties except Wexford in same NUTS III region.

## Answer 7.

    Journey Over 1.5 hr has to be considered the longest commute time. 'tapply' function is used to aggregate it with County column by sum.

```{r}

transport_2011_grouped <- tapply(transport_2011$Journey.One.And.Half.Hours.And.Over, transport_2011$County, sum)
transport_2011_grouped
df <- as.data.frame(sort(tail(sort(transport_2011_grouped),5)))
df <- cbind(rownames(df),df)
names(df)[1] <- 'County'
names(df)[2] <- 'CommuteTimes'
df[,2] <- as.integer(df[,2])

p<-ggplot(data = df, aes(x=df$County, y=df$CommuteTimes)) +
  geom_bar(stat="identity", fill = "#CC79A7") +
  geom_text(aes(label=df$County), vjust=2, size=4, color="white")  +
  theme_minimal() +
  xlab("Top 5 Counties with longest commute times") +
  ylab("Commute Times")
p

```

## Answer 8.

    The proportion is mentioned as the ratio of commute by car containing only single person.

```{r}

occurances <- table(transport_2011$Travelby.Car.Driver)
occurances[names(occurances) == 1]/sum(occurances)

```

    The proportion is `occurances[names(occurances) == 1]/sum(occurances)` which is too small to express it on a visualization.
    

## Answer 9.

    To consider the evidence of substaintial difference, three categories are selected for each three sections: two bounds and one in the middle. The max and min values are taken for each category along with the corresponding 'Electoral Division Region' within County Wexford and juxtaposed together. 

```{r}

agg <- function(df, value) {
  df1 <- tapply(df, transport_2011_wexford$Electoral.Division.Name, sum)
  dfmin <- as.data.frame(head(sort(df1[!is.na(df1)]),1))
  dfmax <- as.data.frame(tail(sort(df1[!is.na(df1)]),1))
  names(dfmin)[1] <- 'min_max'
  names(dfmax)[1] <- 'min_max'
  df2 <- rbind.data.frame(dfmin,dfmax)
  dft <- as.data.frame(c(rep(value,2)))
  names(dft)[1] <- 'type'
  dft <- cbind.data.frame(df2,dft)
  dft <- cbind(rownames(dft),dft)
  names(dft)[1] <- 'Areas'
  return(dft)
}

t1 <- agg(transport_2011_wexford$Travelby.Public.Transport.Comb,'travelby_public')
t2 <- agg(transport_2011_wexford$Travelby.Private.Transport.Comb, 'travelby_private')
t3 <- agg(transport_2011_wexford$Travelby.Soft.Modes.Comb, 'travelby_soft')
l1 <- agg(transport_2011_wexford$Leaving.Before.0630, 'leaving_before_6_30')
l2 <- agg(transport_2011_wexford$Leaving.0801.0830, 'leaving_8_to_8_30')
l3 <- agg(transport_2011_wexford$Leaving.After.0930, 'leaving_after_9_30')
j1 <- agg(transport_2011_wexford$Journey.Under.15.mins, 'journey_under_15min')
j2 <- agg(transport_2011_wexford$Journey.Three.Quarter.Hours.To.Under.One.Hour, 'journey_45min_to_1hr')
j3 <- agg(transport_2011_wexford$Journey.One.And.Half.Hours.And.Over, 'journey_over_1.5hr')

t <- rbind.data.frame(t1,t2,t3)

ggplot(data = t, aes(fill=t$type, y=t$min_max, x=t$Areas)) + 
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=t$Areas), vjust=-1, size=3.5)  +
  scale_fill_viridis(discrete = T, option = "B") +
  ggtitle("Min and Max for Travel Modes") +
  facet_wrap(~t$type) +
  theme_classic() +
  xlab("Areas") +
  ylab("Min and Max Values")

l <- rbind.data.frame(l1,l2,l3)

ggplot(data = l, aes(fill=l$type, y=l$min_max, x=l$Areas)) + 
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=l$Areas), vjust=-1, size=3.5)  +
  scale_fill_viridis(discrete = T, option = "C") +
  ggtitle("Min and Max for Leaving Time") +
  facet_wrap(~l$type) +
  theme_classic() +
  xlab("Areas") +
  ylab("Min and Max Values")

j <- rbind.data.frame(j1,j2,j3)

ggplot(data = j, aes(fill=j$type, y=j$min_max, x=j$Areas)) + 
  geom_bar(position="dodge", stat="identity") +
  geom_text(aes(label=j$Areas), vjust=-1, size=3.5)  +
  scale_fill_viridis(discrete = T, option = "D") +
  ggtitle("Min and Max for Journey Time") +
  facet_wrap(~j$type) +
  theme_classic() +
  xlab("Areas") +
  ylab("Min and Max Values")

```

## Answer 10.
    
    The electoral divisions within 'Southern', 'Eastern and Midlands' and 'North ans West' to be priortized for public transportation are considered by picking highest aggregate mean values. 

```{r}

class(transport_2011$Planning.Region)

for (i in 1:ncol (transport_2011)) 
  if (class (transport_2011[,i]) == "factor") 
    transport_2011[,i] <- as.character(transport_2011[,i])

class(transport_2011$Planning.Region)

transport_2011$Planning.Region[18450:18458] <- "Eastern and Midlands"
transport_2011$Planning.Region[18450:18458]
transport_2011$Planning.Region[transport_2011$Planning.Region == 'South'] <- "Southern"

```

    Before carrying out the operation, there were some missing values in planning region which weren't detected before as the column contains strings as factors and needed to be converted into characters. It is evident from the pattern of data that those missing values are probably 'Eastern and Midlands'.
    
    There is also a single value called 'South' which needed to be converted to 'Southern'.

```{r}
southern <- filter(transport_2011, Planning.Region == 'Southern')
east_mid <- filter(transport_2011, Planning.Region == 'Eastern and Midlands')
north_west <- filter(transport_2011, Planning.Region == 'North and West')

southern_bymean <- tapply(southern$Travelby.Public.Transport.Comb, southern$Electoral.Division.Name, mean)
res1 <- head(sort(southern_bymean, decreasing = TRUE),1)

east_mid_bymean <- tapply(east_mid$Travelby.Public.Transport.Comb, east_mid$Electoral.Division.Name, mean)
res2 <- head(sort(east_mid_bymean, decreasing = TRUE),1)

north_west_bymean <- tapply(north_west$Travelby.Public.Transport.Comb, north_west$Electoral.Division.Name, mean)
res3 <- head(sort(north_west_bymean, decreasing = TRUE),1)

df <- as.data.frame(c(res1,res2,res3))
df <- cbind(rownames(df),df)
names(df)[1] <- 'Electoral_Division'
names(df)[2] <- 'Public_Transportation'
df[,2] <- as.integer(df[,2])

p<-ggplot(data = df, aes(x=df$Electoral_Division, y=df$Public_Transportation)) +
  geom_bar(stat="identity", fill = "#D16103") +
  geom_text(aes(label=df$Electoral_Division), vjust=2, size=4, color="white")  +
  theme_minimal() +
  xlab("Electoral Divisions in Southern, Eastern and Midlands, and North and West respectively") +
  ylab("Total Public Transportation Counts")
p

````